Learning Features and Abstract Actions for Computing Generalized Plans

### 2018-09-06 doc/paper.tex
Blai's last changes

A solution to a generalized problem $\Q$ given the features $F$ and abstract
actions $A_F$ is a partial function $\pi$ that maps \emph{abstract states}
into \emph{abstract actions} such that $\pi$ solves all instances $P$ in $\Q$.
The plan or policy $\pi$ induces a trajectory $s_0,a_0,s_1, \ldots, s_n$ in
an instance $P$ of $\Q$ iff
1)~$s_0$ is the initial state in $P$,
2)~$a_i$ is one of the actions that instantiate $\pi(\bar{s}_i)$ in $s_i$,
3)~$a_i$ is applicable in $s_i$ and $s_{i+1}=f(a_i,s_i)$, and
4)~$s_n$ is a goal state of $P$, $n=\infty$, or $\pi(\bar{s}_n)$ either is
undefined, denotes an abstract action that is not applicable in $\bar{s_n}$,
or no applicable action $a_n$ instantiates $\pi(\bar{s}_n)$.
The policy $\pi$ solves $P$ iff all the induced trajectories reach the
goal of $P$.

@@

\medskip
\noindent The first formulas in $T(\S,\F)$ capture the meaning of $D_1$:

\begin{equation}
  \label{eq:d1}
  D_1(s, t) \ \Leftrightarrow\ \textstyle \bigvee_{f}  selected(f)
\end{equation}

@@

\noindent The second formulas encode the meaning of $D_2$:

\begin{equation}
  \label{eq:d2}
  D_2(s, s', t, t') \ \Leftrightarrow\ \textstyle\bigvee_f  selected(f)
\end{equation}

@@

The third class of formulas relate $D_1$ and $D_2$ by enforcing \emph{soundness}
and \emph{completeness} over  the samples.  Since the abstract actions  are not given,
the qualitative changes in each  transition $(s,s')$ in $\S$
are  taken as templates  of abstract actions. Thus, if $s$ and $t$ are not distinguished by the selected features, 
for each transition $(s, s') \in \S$, there must be a transition $(t, t') \in \S$
such that the two transitions are not distinguished by the selected features either.
This is expressed as

\begin{equation}
  \label{eq:bridge1}
  \neg D_1(s, t) \  \Rightarrow\ \textstyle\bigvee_{t'} \neg D_2(s, s', t, t')
\end{equation}

@@

The theory $T(\S,\F)$ given by formulas (\ref{eq:d1}--\ref{eq:goal})
has a number of atoms that is linear in the size $|{\cal F}|$  of the feature pool $\F$,
and quadratic in the number $m_1$ of expanded states $s$  and
in the number $m_2$ of transitions $(s,s')$ in $\S$.
The number of clauses grows with
$m_1^2 \times (\abs{{\cal F}} + m_2)$, and 
can be roughly cut in half by eliminating symmetries.

We also consider an alternative theory $T_G(\S,\F)$
that trades off completeness guarantees for a much smaller theory size, which
results from not having to consider
all pairs of expanded states or transitions in $\S$.
%
% No need to mark states as well - we just mark transitions, and states are marked implicitly
%
For this, some of the transitions $(s,s')$ in $\S$  are marked as being \emph{goal-relevant},
and completeness is sought \emph{only with respect to these transitions}.
These goal-relevant transitions are selected for being part of a concrete plan
that is computed for each instance $P$ in $\Q$. The theory  $T_G(\S,\F)$ is then as
 $T(\S,\F)$, with the sole difference that the $s$ and $s'$ arguments in the $D_1$ and $D_2$ atoms 
need to be part of some goal-relevant transition.
% This is achieved by
% computing some plans for some instances $P$ in $\Q$, and by adding the
% resulting state transitions to $\S$ at the same time that the first
% arguments $s$ and $(s,s')$ in the $D_1$ and $D_2$ atoms are restricted
% to be marked.
If the set of marked transitions is much smaller than the full sample set, as will typically be the case,
then $T_G(\S,\F)$ will be much smaller than $T(\S,\F)$.
$T_G(\S,\F)$ will guarantee soundness relative to $\S$
but completeness only relative to the the subset of goal-relevant transitions in $\S$. Yet this does not affect
the resulting formal guarantees, which depend only on soundness.

@@

\Omit{
%% commented out
The goal expression $G_F$ can be provided explicitly  in terms of the primitive predicates,
but  more generally can be extracted  from the satisfying assignment $\sigma$ as well.
For this, $G_F = G_{\sigma}$ is defined as the DNF formula whose terms 
correspond to the abstract states over the selected features in $F_{\sigma}$ 
that are false in all expanded states in $\S$ marked as non-goals
and true in some expanded goal state. This DNF formula can   be simplified in standard ways. 
Due to  (\ref{eq:d1}) and (\ref{eq:goal}) that force the selected features in $F_{\sigma}$
to   distinguish goal  from non-goal states, we have that:

@@

The primitive predicates $p$ used in the grammar must be common to all instances $P$ of the generalized problem $\Q$.
They will usually be common domain predicates. Given that parameters $x$ are common as well,
the compositional semantics of DL defines, for any concept $C$ and role $R$ resulting from these rules and 
any state $s$ of $P$, a denotation $C^s$ ($R^s$) \emph{that can be determined from the denotation of the primitive predicates
in $s$ alone}. We will below see examples of derived concepts such as $\exists\, on^* . C_x$,
whose denotation in any Blocksworld state $s$ is the set of blocks that are above $x$ in $s$.
Our numerical features $n_C$ will be defined by the cardinality of these denotations $C^s$.

@@

This means that a policy $\pi$ obtained with a FOND planner from $Q^+_F$ 
solves the generalized problem $\Q$, provided that the abstract actions
learned are sound relative to $\Q$ and that $\Q$ is restricted to the instances $P$ whose
initial and goal conditions comply with   $I_F$ and $G_F$.
However,  for a policy to be correct, the abstract actions do
not have to be sound relative to all reachable states $s$ in the instances $P$ of $\Q$, it  suffices
if they are sound on the states $s$ that are reachable with the policy. We will see an example of this below.

>>>>

A solution to a generalized problem $\Q$ given the features $F$ and abstract
actions $A_F$ is a partial function $\pi$ that maps \emph{abstract states}
into \emph{abstract actions} such that $\pi$ solves all instances $P$ in $\Q$.
The plan or policy $\pi$ induces a trajectory $s_0,a_0,s_1, \ldots, s_n$ in
an instance $P$ of $\Q$ iff
1)~$s_0$ is the initial state in $P$,
2)~$a_i$ is one of the actions that instantiate $\pi(\bar{s}_i)$ in $s_i$,
3)~$a_i$ is applicable in $s_i$ and $s_{i+1}=f(a_i,s_i)$, and
4)~$s_n$ is a goal state of $P$, $n=\infty$, or $\pi(\bar{s}_n)$ is either
undefined, denotes an abstract action that is not applicable in $\bar{s_n}$,
or no applicable action $a_n$ instantiates $\pi(\bar{s}_n)$.
The policy $\pi$ solves $P$ iff all the induced trajectories reach the
goal of $P$.

@@

\medskip
\noindent The first formulas in $T(\S,\F)$ capture the meaning of $D_1$:
%  
\begin{equation}
  \label{eq:d1}
  D_1(s, t) \ \Leftrightarrow\ \textstyle \bigvee_{f}  selected(f)
\end{equation}

@@

The second formulas encode the meaning of $D_2$:
%
\begin{equation}
  \label{eq:d2}
  D_2(s, s', t, t') \ \Leftrightarrow\ \textstyle\bigvee_f  selected(f)
\end{equation}

@@

The third class of formulas relate $D_1$ and $D_2$ by enforcing \emph{soundness}
and \emph{completeness} over  the samples.  Since the abstract actions  are not given,
the qualitative changes in each  transition $(s,s')$ in $\S$
are  taken as templates  of abstract actions. Thus, if $s$ and $t$ are not distinguished by the selected features, 
for each transition $(s, s') \in \S$, there must be a transition $(t, t') \in \S$
such that the two transitions are not distinguished by the selected features either.
This is expressed as
%
\begin{equation}
  \label{eq:bridge1}
  \neg D_1(s, t) \  \Rightarrow\ \textstyle\bigvee_{t'} \neg D_2(s, s', t, t')
\end{equation}

@@

The theory $T(\S,\F)$ given by formulas \eqref{eq:d1}--\eqref{eq:goal}
has a number of atoms that is linear in the size $|{\cal F}|$  of the feature pool $\F$,
and quadratic in the number $m_1$ of expanded states $s$  and
in the number $m_2$ of transitions $(s,s')$ in $\S$.
The number of clauses grows with
$m_1^2 \times (\abs{{\cal F}} + m_2)$, and 
can be roughly cut in half by eliminating symmetries.

We also consider an alternative theory $T_G(\S,\F)$
that trades off completeness guarantees for a much smaller theory size, which
results from not having to consider
all pairs of expanded states or transitions in $\S$.
%
% No need to mark states as well - we just mark transitions, and states are marked implicitly
%
For this, some of the transitions $(s,s')$ in $\S$  are marked as being \emph{goal-relevant},
and completeness is sought \emph{only with respect to these transitions}.
These goal-relevant transitions are selected for being part of a concrete plan
that is computed for each instance $P$ in $\Q$. The theory  $T_G(\S,\F)$ is then as
 $T(\S,\F)$, with the sole difference that the $s$ and $s'$ arguments in the $D_1$ and $D_2$ atoms 
need to be part of some goal-relevant transition.
% This is achieved by
% computing some plans for some instances $P$ in $\Q$, and by adding the
% resulting state transitions to $\S$ at the same time that the first
% arguments $s$ and $(s,s')$ in the $D_1$ and $D_2$ atoms are restricted
% to be marked.
If the set of marked transitions is much smaller than the full sample set, %as will typically be the case,
$T_G(\S,\F)$ is much smaller than $T(\S,\F)$.
$T_G(\S,\F)$ guarantees soundness relative to $\S$
but completeness only relative to the the subset of goal-relevant transitions in $\S$.
Yet this does not affect the resulting formal guarantees, which depend only on soundness.

@@

\Omit{
%% commented out
The goal expression $G_F$ can be provided explicitly  in terms of the primitive predicates,
but  more generally can be extracted  from the satisfying assignment $\sigma$ as well.
For this, $G_F = G_{\sigma}$ is defined as the DNF formula whose terms 
correspond to the abstract states over the selected features in $F_{\sigma}$ 
that are false in all expanded states in $\S$ marked as non-goals
and true in some expanded goal state. This DNF formula can   be simplified in standard ways. 
Due to  \eqref{eq:d1} and \eqref{eq:goal} that force the selected features in $F_{\sigma}$
to   distinguish goal  from non-goal states, we have that:

@@

The primitive predicates $p$ used in the grammar must be common to all instances $P$ of the generalized problem $\Q$.
%They will usually be common domain predicates.
Given that parameters $x$ are common as well,
the compositional semantics of DL defines, for any concept $C$ and role $R$ resulting
from these rules and any state $s$ of $P$, a denotation $C^s$ ($R^s$)
\emph{that can be determined from the denotation of the primitive predicates in $s$ alone}.
Below we see examples of derived concepts such as $\exists\, on^* . C_x$, whose denotation
in any Blocksworld state $s$ is the set of blocks that are above $x$ in $s$.
The numerical features $n_C$ are defined by the cardinality of such denotations $C^s$.

@@

This means that a policy $\pi$ obtained with a FOND planner from $Q^+_F$ 
solves the generalized problem $\Q$, provided that the abstract actions
learned are sound relative to $\Q$ and that $\Q$ is restricted to the instances $P$ whose
initial and goal conditions comply with   $I_F$ and $G_F$.
However,  for a policy to be correct, the abstract actions do
not have to be sound relative to all reachable states $s$ in the instances $P$ of $\Q$, it  suffices
if they are sound on the states $s$ that are reachable with the policy. We see an example of this below.
